Estimating duration in the presence of misclassification

### 2024-03-08 biology-data.tex
Overleaf comments / two parts intro

The Ct value decreases as the amount of DNA present in the original sample increases~\autocite{hanRTPCR}.
If there is more RNA present, more DNA is created, and hence fewer cycles are required to reach detection.
As the number of DNA copies doubles each cycle, the Ct value decreases increases linearly on the log-scale each cycle (see \cref{biology-data:fig:ct-calibration}).
A \emph{standard curve} translates the Ct value to viral load for a given test.
They are produced by measuring the Ct value using reference samples with known viral load.
As seen in \cref{biology-data:fig:ct-calibration}, they vary between studies.
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{biology-data/ct-calibration}
    \caption[Example standard Ct curves]{%
        Example standard curves, which relate the logarithm of the viral copy number (the number of copies of the target DNA or RNA in the sample) to the measured Ct value.
        Reproduced from \textcite{hanRTPCR} with permission.
    }
    \label{biology-data:fig:ct-calibration}
\end{figure}

@@

An important concept throughout this thesis will be that of \emph{detectable} individuals.
An individual is defined as detectable between the time they start testing positive on a RT-PCR test and the time they stop testing positive on a RT-PCR test.
It is not possible to directly observe whether an individual is detectable or not due to noise in the RT-PCR test (see \cref{biology-data:sec:PCR}).
However, viewing the number of RNA genome copies in an individual as a latent process this can then be defined as the time when the number of RNA genome copies is above the limit of detection of the RT-PCR test (defined in \cref{biology-data:sec:measurement-error}).
While this definition is not perfectly precise, it is clear enough for the purposes it used in this thesis.

@@

False positives are very rare for SARS-CoV-2 RT-PCR tests.
In summer 2020, the CIS performed 208,730 PCR tests and found only 159 positives~\autocite[section 5]{cisMethodsONS}.
If these were all false positives, then the test specificity would be 99.9\%.
In reality, it is almost certain that some of these are true positives, and hence this is a lower-bound on the true test senstivity.

RT-PCR enables the detection of SARS-CoV-2 viral RNA in a sample.
However, this does not necessarily mean that the virus is viable (\ie able to reproduce)~\autocite{puhachSARSCoV2}.
High viral loads (low Ct values) are more likely to be viable, but more sophisticated tests are required to determine viability.
The gold standard procedure is attempting to grow and recover viral material from a sample, a process known as \emph{cell culturing}~\autocite{singanayagamDuration,puhachSARSCoV2,hakkiOnset}.
Further details on cell culturing are beyond the scope of this thesis.

>>>>

The Ct value decreases as the amount of DNA present in the original sample increases~\autocite{hanRTPCR}.
If there is more RNA present, more DNA is created, and hence fewer cycles are required to reach detection.
As the number of DNA copies doubles each cycle, the Ct value decreases increases linearly on the log-scale each cycle (see \cref{biology-data:fig:ct-calibration}).
A \emph{standard curve} translates the Ct value to viral load for a given test.
A standard curve is produced by measuring the Ct value using reference samples with known viral load, and, as seen in \cref{biology-data:fig:ct-calibration}, they vary between studies.
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{biology-data/ct-calibration}
    \caption[Example standard Ct curves]{%
        Example standard curves, which relate the logarithm of the viral copy number (the number of copies of the target DNA or RNA in the sample) to the measured Ct value.
        Reproduced from \textcite{hanRTPCR} with permission.
    }
    \label{biology-data:fig:ct-calibration}
\end{figure}

@@

An important concept throughout this thesis will be that of \emph{detectable} individuals.
An individual is defined as detectable between the time they would first test positive on a perfect RT-PCR test (if tested) and the time they would stop doing so.
It is not possible to conclusively determine whether an individual is detectable because perfect RT-PCR tests do not exist (see \cref{biology-data:sec:PCR}).
However, viewing the number of RNA genome copies in an individual as a latent process this can then be defined as the time when the number of RNA genome copies is above the limit of detection of the RT-PCR test (defined in \cref{biology-data:sec:measurement-error}).
While this definition is not perfectly precise, it is clear enough for the purposes it used in this thesis.

@@

False positives are very rare for SARS-CoV-2 RT-PCR tests.
In summer 2020, the CIS performed 208,730 PCR tests and found only 159 positives~\autocite[section 5]{cisMethodsONS}.
If these were all false positives, then the test specificity would be 99.9\%.
In reality, it is almost certain that some of these are true positives, and hence this is a lower-bound on the true test specificity.

RT-PCR enables the detection of SARS-CoV-2 viral RNA in a sample.
However, this does not necessarily mean that the virus is viable (\ie able to reproduce, a necessary condition for transmission to occur)~\autocite{puhachSARSCoV2}.
High viral loads (low Ct values) are more likely to be viable, but more sophisticated tests are required to determine viability.
The gold standard procedure is attempting to grow and recover viral material from a sample, a process known as \emph{cell culturing}~\autocite{singanayagamDuration,puhachSARSCoV2,hakkiOnset}.
Further details on cell culturing are beyond the scope of this thesis.

### 2024-03-08 introduction-comments.tex
Overleaf comments / two parts intro

Various forms of \emph{reproduction number} were prominent throughout the COVID-19 pandemic~\autocite{pellisEstimation}.
Broadly, a reproduction number describes the number of secondary infections generated by a single infection.
Here, following \textcite{pellisEstimation},  building on over a century of infectious disease dynamics literature, I define three that are broadly useful.

The most famous reproduction number is the \emph{basic reproduction number}, $\R$.
$\R$ is the mean number of secondary infections generated by a single infection in a \emph{fully susceptible} population with \emph{standard behaviour patterns}.
A fully susceptible population means none of its members have any immunity to the disease.
Note that this definition is a constant for a given pathogen and population.
In the early stages of an epidemic caused by a novel pathogen, the number of individuals with immunity is negligible, and hence $\R$ describes this part of the epidemic well.
$\R$ has a threshold value of 1: if $\R < 1$ then the epidemic will die out.

The \emph{effective reproduction number}, $\Re$, is the mean number of secondary infections generated by a single infection at time $t$.
In particular, it uses the behaviour and immunity status at time $t$.
With constant behaviour and a disease that grants immunity following infection, $\Re$ declines smoothly.
$\Re$ describes the current dynamics of the epidemic: $\Re > 1$ means the epidemic is growing, $\Re < 1$ means it is shrinking, and $\Re = 1$ means it is stable.
In simple epidemic models (see \cref{E-SEIR:sec:SIR}), $\Re$ is proportional to the product of the rate of contacts, the riskiness of those contacts, and the proportion of the population that is susceptible.
Therefore, it informs the strength of intervention, \eg social distancing or vaccination, required to control the epidemic.

@@

These definitions ignore many complexities that arise when theoretically defining or practically estimating any of the reproduction numbers~\autocite{pellisEstimation}.
For example, effects due to changing weather, what a ``mean infection'' means in a heterogeneous population, and the lack of a ground truth for these numbers in almost all practical situations.
However, their intuitive simplicity and convenient properties in simple epidemic models (see \cref{E-SEIR:sec:mechanistic-models}) mean that they are useful concepts for guiding policy.

Reproduction numbers are strongly connected to \emph{growth rates}.
Over short periods of time, epidemics tend to grow exponentially, and hence have an exponential growth rate, normally denoted $r$.
For a given disease and population, an analogous growth rate can be defined for each reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
The advantage of using the growth rate is that they are easier to estimate and capture how quickly spread happens in calendar time; however, they do not capture the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.

@@

However, incidence and transmission are almost never directly observed.
Therefore, they need to be estimated from other data sources.

\emph{Prevalence} is the proportion of the population with the disease at a given time.
Prevalence can be observed, most commonly with a prevalence survey.

See \cref{E-inc-prev:sec:definitions} for a formal definition of incidence and prevalence.
Prevalence and incidence are strongly related, with the relationship determined by how long the disease lasts (see \cref{E-inc-prev}).

@@

\section{Data sources} \label{intro:sec:data-sources}

A wide variety of designed studies, administrative data, and other data sources have been used for pandemic and epidemic surveillance.
In the context of the SARS-CoV-2 pandemic in England, the most prominent were: community testing (pillar 2), the number of severe events (\eg hospital admissions or deaths), seroprevalence testing (seroprevalence is the proportion of people with antibodies to a disease, a marker of prior infection), and prevalence surveys.
Each of these data sources have strengths and weaknesses, see \textcite{royalSocietyRnumber} for further details.

Community testing is broadly available, but does not test a random sample of the population.
The number of tests conducted varies over time and space, for a variety of known and unknown reasons.
These include: policy on who is eligible for testing, the availability of tests, and the willingness of individuals to be tested.
Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
For large parts of the pandemic, community testing was widely available in the UK.
Therefore, this is a large dataset, allowing detailed analyses to be performed, including stratification by small geographic areas or age groups.

Seroprevalence testing provides information on a marker of recent prior infection, by measuring ..????
Prior infection is more common than current infection, hence seroprevalence estimation requires a smaller sample size than current prevalence.
However, there is a delay between infection and the development of antibodies, hence seroprevalence is not a real-time measure of the burden of the disease.
Furthermore, seroprevalence data is normally based on convenience samples (\eg blood donors) who are not representative of the general population.

Severe events are generally a reliable measure of the burden of the disease.
The criteria for being a severe event are normally well-defined and generally well-recorded in the UK context (reference?).
Therefore, the data is subject to fewer biases than community testing.
However, SARS-CoV-2 rarely leads to severe events in younger individuals, meaning severe events only capture sequelae of transmission in a subset of the population.

Prevalence surveys in representative samples of the population are the most reliable measure of the burden of the disease.
They are not subject to the biases of community testing and observe the whole population, unlike severe events.
However, they are expensive to power to inform detailed analyses.
Often, some modelling is required to borrow information across strata or time to provide stable estimates (see \cref{E-intro:sec:cis}).

The use of prevalence surveys for estimating incidence and transmission is the focus of this thesis.
In particular, I focus on the CIS (Coronavirus (COVID-19) Infection Survey), run by the Office for National Statistics (ONS)~\autocite{CIS,cisMethodsONS}.
The CIS was a longitudinal study, testing a large number of indidividuals weekly or monthly for SARS-CoV-2; it is described in detail in \cref{E-intro:sec:cis}.
Early in the pandemic, a methodology was developed for using CIS data to estimate prevalence (see \cref{E-intro:sec:cis}).

A second prevalence survey ran in the UK, known as REACT (REal-time Assessment of Community Transmission)~\autocite{rileyResurgence,rileyREACT}.
REACT was a series of cross-sectional surveys, run across 19 rounds between May 2020 and March 2022.
Each round consisted of at least 100,000 randomly-sampled individuals from the UK contributing swabs for RT-PCR testing and questionnaire data~\autocite{elliottTwin}.

@@

The statistical framework developed for HIV is applicable to SARS-CoV-2.
(see \cref{E-inc-prev:sec:infection-process}).

However, Unlike HIV/AIDS, recovery from SARS-CoV-2 infection is possible, so recoveries need to be considered in the backcalculation.
This introduces a dependency structure in the prevalence data (see \cref{E-inc-prev:sec:prevalence-process}).
The development of this framework makes clear that the duration of positivity (how long an individual tests positive for SARS-CoV-2) is the key quantity relating prevalence to incidence.
Therefore, estimation of duration is central to this thesis (see \cref{intro:sec:duration}).

Furthermore, prevalence surveys are noisy because they are based on population sampling.
This observation noise needs to be considered in the development of the backcalculation approach (see \cref{E-inc-prev:sec:observation-process}).

The original formulations of backcalculation neglected any observation noise.
AIDS is almost always diagnosed and recorded in developed countries.
That is, the observations of AIDS diagnoses are the same as the true incidence of AIDS.
An exception is that the reporting delay may be important~\autocite{paganoHIV}; but reporting delay is not important for this thesis's work.
Since its development, backcalculation has been further developed for a wide-variety of contexts.
% Reporting delay is not relevant to the studies in this thesis, and I ignore it in this and further discussion.

\Textcite{abbottCISincidence} applied backcalculation to CIS data to estimate the incidence of SARS-CoV-2 in the UK.
They used a Gaussian process prior to smooth the incidence, applying the prior to either smooth the log-incidence or its derivative; smoothing of the incidence is required to stabalise the estimates~\autocite{brookmeyerBackcalculation}.
\Textcite{abbottCISincidence} used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing to inform the backcalculation, which are based on a longitudinal study in healthcare workers (see \cref{intro:sec:duration-SARS-CoV-2}).
They fit to summary statistics of the CIS data, using a daily, independent Gaussian likelihood with parameters estimated from the posterior point estimate and 95\% credible interval published by the ONS.

@@

Each week, the members of SPI-M would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
The nowcasting estimates were mainly based on mechanistic models.
These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value~\autocite{parkCombining}.
The consensus value was published on \url{gov.uk} and informed the government's response to the epidemic~\autocite{govRnumber}.
The academic groups contributing estimates used a variety of models, reviewed by \textcite{royalSocietyRnumber} (see appendix 2 of that review for details of the specific implementations).
Most compartmental models used by SPI-M-O were age-stratified, as differences between age groups are important for both the transmission and severity of SARS-CoV-2 (see \cref{E-SEIR:sec:structured-populations}).

In this thesis, I focus on the use of data from prevalence surveys.
Several compartmental models incorporated data from SARS-CoV-2 prevalence surveys, alongside other data sources~\autocite{daviesAssociation,ironsEstimating,knockKey,nicholsonImproving,pooleyEstimation,birrellRTM2}.
These models incorporate prevalence survey data by adding additional compartments representing RT-PCR individuals, who may not be infectious (this is common, see \cref{E-biology-data}); except \textcite{nicholsonImproving} who have a more sophisticated approach (below) and \textcite{ironsEstimating}, who assume RT-PCR positivity coincides with infectiousness.
The length of stay in these compartments (the duration of time positive) is assumed to follow a gamma distribution, with the parameters estimated from external data or having at most one free parameter.
The proportion of the population in these compartments are then fit to the prevalence survey data using a binomial likelihood where possible, although some approaches take a Gaussian approximation.
These approaches are similar to how these models are fit to other data sources, such as hospital admissions or deaths.

@@

\subsection{Methodology}

Two approaches are commonly used to estimate the duration of a disease: either modelling an underlying biomarker or modelling the duration directly~\autocite{sweetingEstimating}.
For SARS-CoV-2, both approaches would be based on the results of RT-PCR testing (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}\todo{move to where RT-PCR / PCR is first mentioned}).

@@

The overall question that I investigate in this thesis is: how can prevalence surveys best be used to understand incidence and transmission?
Prevalence surveys provide data that is far more representative of the general population than other data sources (see \cref{intro:sec:data-sources}), therefore, conclusions based on them should be far more robust.
There are two components to this question.
First, how can prevalence surveys be designed to help estimate these quantities.
Second, what statistical methodology can be used to estimate these quantities from prevalence surveys.
This thesis will focus on the second component.
However, natural extensions of the work I undertake would inform the first component.

@@

First, in \cref{E-inc-prev:sec:observation-process}, I provide a theoretic justification for the use of deterministic backcalculation in this setting.
A deterministic backcalculation approach was used by \textcite{abbottCISincidence} and I will use it in \cref{E-backcalc}.
This approach models the prevalence at any time as a deterministic function of the incidence, rather than as a random variable.
I show that, this is a good approximation as long as the proportion of the population sampled is small.
In particular, it does not require the prevalence or incidence to be large.

>>>>

Various forms of \emph{reproduction number} were prominent throughout the SARS-CoV-2 pandemic~\autocite{pellisEstimation}.
Broadly, a reproduction number describes the number of secondary infections generated by a single infection.
Here, following \textcite{pellisEstimation},  building on over a century of infectious disease dynamics literature, I define three that are broadly useful.

The most famous reproduction number is the \emph{basic reproduction number}, $\R$.
$\R$ is the mean number of secondary infections generated by a single infection in a \emph{fully susceptible} population with \emph{standard behaviour patterns}.
A fully susceptible population means none of its members have any immunity to the disease.
Note that this definition implies $\R$ is constant for a given pathogen and population.
In the early stages of an epidemic caused by a novel pathogen, the number of individuals with immunity is negligible, and hence $\R$ describes this part of the epidemic well.
$\R$ has a threshold value of 1: if $\R < 1$ then the epidemic will die out.

The \emph{effective reproduction number}, $\Re$, is the mean number of secondary infections generated by a single infection at time $t$.
In particular, it uses the behaviour and immunity status of the population at time $t$.
With constant behaviour and a disease that grants immunity following infection, $\Re$ declines smoothly.
$\Re$ describes the current dynamics of the epidemic: $\Re > 1$ means the epidemic is growing, $\Re < 1$ means it is shrinking, and $\Re = 1$ means it is stable.
In simple epidemic models (see \cref{E-SEIR:sec:SIR}), $\Re$ is proportional to the product of the rate of contacts, the riskiness of those contacts, and the proportion of the population that is susceptible.
Therefore, it informs the strength of intervention, \eg social distancing or vaccination, required to control the epidemic.

@@

These definitions ignore many complexities that arise when theoretically defining or practically estimating any of the reproduction numbers~\autocite{pellisEstimation}.
For example, effects due to seasonality, what a ``mean infection'' means in a heterogeneous population, and the lack of a ground truth for these numbers in almost all practical situations.
However, their intuitive simplicity and convenient properties in simple epidemic models (see \cref{E-SEIR:sec:mechanistic-models}) mean that they are useful concepts for guiding policy.

Reproduction numbers are strongly connected to \emph{growth rates}.
Over short periods of time, epidemics tend to grow (or shrink) exponentially, and hence have an exponential growth rate, normally denoted $r$.
For a given disease and population, an analogous growth rate can be defined for each reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
The advantage of using the growth rate is that they are easier to estimate and capture how quickly spread happens in calendar time; however, they do not capture the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.

@@

\emph{Prevalence} is the proportion of the population with the disease at a given time.
Whether an individual is \emph{prevalent} (has the disease) is normally measurable.
Therefore, the prevalence of the population can be estimated with a prevalence survey.
However, these are very expensive to power appropriately; SARS-CoV-2 is unusual in that its impact is so large that the UK government did this.
\cref{E-inc-prev:sec:definitions} provides a formal definition of incidence and prevalence.
Prevalence and incidence are strongly related, with the relationship determined by by the duration of an infection (see \cref{E-inc-prev}).

@@

However, incidence and transmission and prevalence, are almost never directly observed.
Therefore, they need to be estimated from other data sources.

Introduce next section eg incidence / transmission not observed directly, need infer from data such as next section.

\section{Data sources in SARS-CoV-2} \label{intro:sec:data-sources}

A wide variety of designed studies, administrative data, and other data sources have been used for pandemic and epidemic surveillance.
In the context of the SARS-CoV-2 pandemic in England, the most prominent were: community testing (known as pillar 2 testing), the number of severe events (\eg hospital admissions or deaths), seroprevalence testing (seroprevalence is the proportion of people with antibodies to a disease, a marker of prior infection), and prevalence surveys.
Each of these data sources have strengths and weaknesses, see \textcite{royalSocietyRnumber} for further details.

Community testing is broadly available, but does not test a random sample of the population.
The number of tests conducted varies over time and space, for a variety of known and unknown reasons.
These include: policy on who is eligible for testing, the availability of tests, and the willingness of individuals to be tested.
Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
For large parts of the pandemic, community testing was widely available in the UK.
Therefore, there are many test results, allowing detailed analyses to be performed, including stratification by small geographic areas or age groups.

Seroprevalence testing provides information on a marker of recent prior infection.
Prior infection is more common than current infection, hence seroprevalence estimation requires a smaller sample size than current prevalence.
However, there is a delay between infection and the development of antibodies, hence seroprevalence is not a real-time measure of the burden of the disease.
Furthermore, seroprevalence data is normally based on convenience samples (\eg blood donors) who are not representative of the general population.

Severe events counts are generally a reliable measure of the burden of the disease.
The criteria for being a severe event are normally well-defined and generally well-recorded in the UK context\todo{ref severe events criteria}.
Therefore, these data are subject to fewer biases than community testing.
However, SARS-CoV-2 rarely leads to severe events in younger individuals, meaning severe event counts are insensitive to changes in transmission only occurring in younger individuals.

Prevalence surveys in representative samples of the population are the most reliable measure of the burden of the disease.
They are not subject to the biases of community testing and observe the whole population, unlike severe events, although they can suffer from non-response bias.
However, they are expensive to power to inform detailed analyses.
Often, some modelling is required to borrow information across strata or time to provide stable estimates, as well as to correct for non-response bias (see \cref{E-intro:sec:cis}).

The use of prevalence surveys for estimating incidence and transmission is the focus of this thesis.
In particular, I focus on the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
The CIS was a longitudinal study, testing a large number of indidividuals weekly or monthly for SARS-CoV-2; it is described in detail in \cref{E-intro:sec:cis}.
Early in the pandemic, a methodology was developed for using CIS data to estimate prevalence (see \cref{E-intro:sec:cis}).

A second prevalence survey ran in the UK, known as REACT (REal-time Assessment of Community Transmission)~\autocite{rileyResurgence,rileyREACT}.
REACT was a series of cross-sectional surveys, run across 19 rounds between May 2020 and March 2022.
Each round consisted of at least 100,000 randomly-sampled individuals from the UK contributing swabs for testing and questionnaire data~\autocite{elliottTwin}.

@@

The statistical framework developed for HIV is applicable to SARS-CoV-2.
(see \cref{E-inc-prev:sec:infection-process}).
However, unlike with HIV/AIDS, recovery from SARS-CoV-2 infection is possible, so recoveries need to be considered in the backcalculation.
This introduces a dependency structure in the prevalence data (see \cref{E-inc-prev:sec:prevalence-process}).
The development of this framework makes clear that the duration of positivity (how long an individual tests positive for SARS-CoV-2) is the key quantity relating prevalence to incidence.
Therefore, estimation of duration is central to this thesis (see \cref{intro:sec:duration}).

Furthermore, prevalence surveys are noisy because they are based on population sampling.
The original formulations of backcalculation neglected any observation noise.
AIDS is almost always diagnosed and recorded in developed countries.
That is, the observations of AIDS diagnoses are the same as the true incidence of AIDS.
An exception is that the reporting delay may be important~\autocite{paganoHIV}; but reporting delay is not important for this thesis's work.
Since its development, backcalculation has been further developed for a wide-variety of contexts.
I incorporate observation noise into the backcalculation approach in \cref{inc-prev:sec:observation-process}.
% Reporting delay is not relevant to the studies in this thesis, and I ignore it in this and further discussion.

\Textcite{abbottCISincidence} applied backcalculation to CIS data to estimate the incidence of SARS-CoV-2 in the UK.
They used a Gaussian process prior to smooth the incidence, applying the prior to either smooth the log-incidence or its derivative; smoothing of the incidence is required to stabilise the estimates~\autocite{brookmeyerBackcalculation}.
\Textcite{abbottCISincidence} used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing to inform the backcalculation, which are based on a longitudinal study in healthcare workers (see \cref{intro:sec:duration-SARS-CoV-2}).
They fit to summary statistics of the CIS data, using a daily, independent Gaussian likelihood with parameters estimated from the posterior point estimate and 95\% credible interval published by the ONS.

@@

Each week, the members of SPI-M-O would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
The nowcasting estimates were mainly based on mechanistic models.
These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value~\autocite{parkCombining}.
The consensus value was published on \url{gov.uk} and informed the government's response to the epidemic~\autocite{govRnumber}.
The academic groups contributing estimates used a variety of models, reviewed by \textcite{royalSocietyRnumber} (see appendix 2 of that review for details of the specific implementations).
Most compartmental models used by SPI-M-O were age-stratified, as differences between age groups are important for both the transmission and severity of SARS-CoV-2 (see \cref{E-SEIR:sec:structured-populations}).

In this thesis, I focus on the use of data from prevalence surveys.
Specifically, surveys which collect swabs from individuals for testing by RT-PCR (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}).
Several compartmental models incorporated data from SARS-CoV-2 prevalence surveys, alongside other data sources~\autocite{daviesAssociation,ironsEstimating,knockKey,nicholsonImproving,pooleyEstimation,birrellRTM2}.
These models incorporate prevalence survey data by adding additional compartments representing RT-PCR positive individuals, who may not be infectious (this is common, see \cref{E-biology-data}); except \textcite{nicholsonImproving} who have a more sophisticated approach (below) and \textcite{ironsEstimating}, who assume RT-PCR positivity coincides with infectiousness.
The length of stay in these compartments (the duration of time positive) is assumed to follow a gamma distribution, with the parameters estimated from external data or having at most one free parameter.
The proportion of the population in these compartments are then fit to the prevalence survey data using a binomial likelihood where possible, although some approaches take a Gaussian approximation.
These approaches are similar to how these models are fit to other data sources, such as hospital admissions or deaths.

@@

\subsection{Approaches}

Two approaches are commonly used to estimate the duration of a disease: either modelling an underlying biomarker or modelling the duration directly~\autocite{sweetingEstimating}.
For SARS-CoV-2, both approaches would be based on the results of RT-PCR testing.

@@

The overall question that I investigate in this thesis is: can prevalence surveys alone be used to infer incidence and transmission, and if so how?
Prevalence surveys provide data that is far more representative of the general population than other data sources (see \cref{intro:sec:data-sources}), therefore, conclusions based on them should be far more robust.
There are two components to this question.
First, how can prevalence surveys be designed to help estimate these quantities.
Second, what statistical methodology can be used to estimate these quantities from prevalence surveys.
This thesis will focus on the second component.
However, natural extensions of the work I undertake would inform the first component.

@@

First, in \cref{E-inc-prev:sec:observation-process}, I provide a theoretic justification for the use of deterministic backcalculation in this setting.
A \emph{deterministic} backcalculation approach was used by \textcite{abbottCISincidence} and I will use it in \cref{E-backcalc}.
The deterministic approach models the prevalence at any time as a deterministic function of the incidence, rather than as a random variable.
I show that, this is a good approximation as long as the proportion of the population sampled is small.
In particular, it does not require the prevalence or incidence to be large.

### 2024-03-08 transmission.tex
Overleaf comments / two parts intro

An alternative would be an inference procedure that estimates incidence proportion, fit directly to the RT-PCR test results.
Denote the quantity to be inferred (which could be either incidence proportion or prevalence) at time $t$ in strata $i$ as $\pi_{it}$.
The strata will represent at least age group and region, although including ethnicity and sex is also due to non-response in the survey (see \cref{E-intro:sec:cis-strata}).
For many combinations of stratum $i$ and day $t$, the number of positives will be small, often fewer than 5; therefore some form of borrowing between time and strata is required.
The statistical model and inference procedure should have the following properties.
\begin{enumerate}
    \item The smoothing model over time should penalise changes from continuous exponential growth.
        Penalising changes from continuous exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
        Furthermore, such penalisation has previously been shown to improve estimates of SARS-CoV-2 prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
    \item Limit estimates to epidemiologically and biologically plausible values.
        That is, incidence proportion and prevalence should be in the interval $[0, 1]$.
        % While incidence proportion can theoretically be greater than 1, this would imply that some individuals were infected multiple times in a day which is not biologically plausible.
    \item Have sufficient parameters to incorporate all necessary covariates to correct for non-response bias (see \cref{E-biology-data:sec:cis}).
    \item Borrow information across the different covariates.
        The borrowing should penalise away from changes in the distribution of incidence/prevalence across strata in the same geographic region being constant.
        Standard stratified, mechanistic transmission model (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) predict this, if behavioural changes are independent of strata and there is negligible impact of immunity.
    \item Be computationally efficient.
        The model will be fit to granular data and therefore needs to be run within the SRS, with its computational limitations (see \cref{E-biology-data:sec:SRS}).
\end{enumerate}

@@

I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using the methodology of \textcite{vanleeuwenTime,vanleeuwenAugmenting}.
%; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocite{vanleeuwenTime,birrellRealtime}.
% These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
This method starts with pre-pandemic baseline estimates of $\kappa_{aa'}$ from the POLYMOD study~\autocite{mossongSocial}, and time use from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on each contact's age and location (home, work, school, leisure, transport, or other).
UKTUS asked a representative sample of the UK population about how much time they spent doing various activities (\eg shopping and spending time at home) each day.
\Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
\Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
Google mobility data are based on the location of Android users over time, relative to a pre-pandemic baseline.
The mobility and school data are averaged over a week then used to down or up-weight the number of contacts in each setting, before combining contacts across settings to form $\kappa_{aa'}(t)$.
$\kappa_{aa'}(t)$ is assumed to be piecewise constant, changing only at the start of each week.
For further details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.

>>>>

An alternative approach would be to estimate incidence proportion directly from the RT-PCR test results.

Denote the quantity to be inferred (which could be either incidence proportion or prevalence) at time $t$ in strata $i$ as $\pi_{it}$.
The strata will represent at least age group and region, although including ethnicity and sex is also included due to non-response in the survey (see \cref{E-intro:sec:cis-strata}).
For many combinations of stratum $i$ and day $t$, the number of positives will be small, often fewer than 5; therefore some form of borrowing between time and strata is required.
The statistical model and inference procedure should have the following properties.
\begin{enumerate}
    \item The smoothing model over time should penalise changes from continuous exponential growth.
        Penalising changes from continuous exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
        Furthermore, such penalisation has previously been shown to improve estimates of SARS-CoV-2 prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
    \item Limit estimates to epidemiologically and biologically plausible values.
        That is, incidence proportion and prevalence should be in the interval $[0, 1]$.
        % While incidence proportion can theoretically be greater than 1, this would imply that some individuals were infected multiple times in a day which is not biologically plausible.
    \item Have sufficient parameters to incorporate all necessary covariates to correct for non-response bias (see \cref{E-biology-data:sec:cis}).
    \item Borrow information across the different covariates.
        The borrowing should penalise away from changes in the distribution of incidence/prevalence across strata in the same geographic region being constant.
        Standard stratified, mechanistic transmission model (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) predict this, if behavioural changes are independent of strata and there is negligible impact of immunity.
    \item Be computationally efficient.
        The model will be fit to granular data and therefore needs to be run within the SRS, with its computational limitations (see \cref{E-biology-data:sec:SRS}).
\end{enumerate}

@@

I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using the methodology of \textcite{vanleeuwenTime,vanleeuwenAugmenting}.
%; these estimates have been used for previous modelling of the SARS-CoV-2 pandemic~\autocite{vanleeuwenTime,birrellRealtime}.
% These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
This method starts with pre-pandemic baseline estimates of $\kappa_{aa'}$ from the POLYMOD study~\autocite{mossongSocial}, and time use from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on each contact's age and location (home, work, school, leisure, transport, or other).
UKTUS asked a representative sample of the UK population about how much time they spent doing various activities (\eg shopping and spending time at home) each day.
\Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
\Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
Google mobility data are based on the location of Android users over time, relative to a pre-pandemic baseline.
The mobility and school data are averaged over a week then used to down or up-weight the number of contacts in each setting, before combining contacts across settings to form $\kappa_{aa'}(t)$.
$\kappa_{aa'}(t)$ is assumed to be piecewise constant, changing only at the start of each week.
For further details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.
